On mapping errors and seeding heuristics for DNA alignment

### 2019-04-03 MEM.tex
First version ready to submit full manuscript (not copy-edited).

\title{Estimating mapping quality when using seeding heuristics for
DNA alignment}

@@

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies demands faster
algorithms to analyze sequencing data. The fate of most reads is to be
mapped to a reference sequence, typically a genome. Modern mappers rely on
heuristics to gain speed at a reasonable cost for accuracy. In the seeding
approach, short matches between the reads and the genome are used to
narrow the search to a set of candidate locations. Several seeding
variants used in modern mappers show good empirical performance but
theoretical results are still lacking, making them difficult to calibrate
or top optimize.
% Summary of the results.
Here we develop a theory to estimate the probability that reads are
mismapped due to a seeding failure. We describe the properties of simple
exact seeds, skip-seeds used in Bowtie2 and MEM (Maximal Exact Match)
seeds used in BWA-MEM. 
% Innovation.
The major innovation is to use concepts from analytic combinatorics to
represent reads as abstract sequences, from which we estimate the
probabilities of interest.
% Durability.
We provide several algorithms, which combined together give a practicial
solution to the problem of calibrating seeding heuristic for short reads.
% Scope.
These results significantly improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
\end{abstract}

@@

Suppose that a DNA fragment drawn at random from a known
genome\footnote{We will assume that the sequence of the genome is known
exactly and with absolute certainty. This is not the case in practical
application, but this assumption will allow us to develop a theory
centered on the read.} is sequenced with an imperfect instrument; how can
we identify the location of the fragment in the genome?

We will refer to this question as the \emph{true location problem}.
It has applications in countless domains, like discovering
disease-causing mutations; diagnosing viral, bacterial or fungal
infections; diagnosing cancer and choosing appropriate therapies;
selecting breeds of interest in agriculture; tracing human migrations from
ancient bones; identifying the donor of a biological sample; diagnosing
genetic diseases at birth; finding compatible graft donors \textit{etc}.

At first sight, whether a read read can be mapepd to its correct location
seems to depend only on its length and on the error rate of the sequencer.
However, there is another important factor to consider: most genomes
contain repeated sequences, \textit{i.e.} relatively large subsequences
that are present at more than one location. For instance, we cannot map
the fragment with certainty if there exists an exact copy of the sequence
somewhere else in the genome, because we cannot know which of the two
copies was sequenced.

This case is rare though. Repeated sequences are in general not identical,
but merely homologous\footnote{By that we mean that drawing nucleotides
uniformly at random is very unlikely to produce sequences with the
observed degree of identity, meaning that the similarity is due some
biological duplication process.}. So when the DNA fragment is duplicated,
we can still map it, as long as we can distinguish it from its duplicates.
This in turn depends on their similarity with the target fragment and on
the error rate of the sequencer.

@@

We will refer to this task as the \emph{best location problem}. It amounts
to finding the optimal alignment between two sequences, and for this
reason has received substantial attention in bioinformatics. For
simplicity we will assume that the true location is always the best. We
will relax this assumption in section section~\ref{sec:not_the_best}, but
until then we will consider that the solution of the true location problem
is the best location.

@@

The slowness of the alignment algorithm imposes a trade-off between speed
and accuracy: If the seeding step is set to produce many candidates, it is
likely to retrieve the candidate. The downside is that this will take
long. Conversely, if the seeding step is set to produce few candidates,
the process will run faster but the target is less likely to be in the
candidate list.

@@

The distinction is important because duplicates of the target are similar
to the read (due to their similarity to the target), so a failure of the
first kind is difficult to distinguish from a success. A failure of the
second kind is easier to recognize because in this case the best hit is
not similar to the read.

@@

If a seed-based mapping algorithm returns an incorrect location, then it
must be the case that the best location was never tested in the second
phase, otherwise it would have been singled out by the exact alignment
algorithm.

This observation is fundamental: in order to estimate the mismapping rate
of an algorithm, we must know how often the associated seeding heuristic
is off target. For lack of theoretical framework, this information is
missing for nearly all existing seeding methods, meaning that current
mappers cannot be properly calibrated.

The focus of this article is to develop a theory to estimate the
probability that seeding is off target under the most popular seeding
heuristics for short reads. Our previous work pioneered a method to
compute seeding probabilities but it did not distinguish off-target from
null seeding~\cite{filion2017analytic,filion2018analytic}, and therefore
did not provide a way to control the mismapping rate. Other authors
investigated the reliability of mapping algorithms~\cite{pmid23872968},
but they focused on the probability that random sequences have a single
hit, recognizing that solving the problem requires taking into
consideration the repeat structure of the genome. Here we address this
problem in general terms and we provide a practical way to estimate the
mismapping rate in complex genomes.

The rest of this document is organized to present the content in a
didactic manner. The order of the examples is meant to facilitate the
aquisition of the main concepts, meaning that we sometimes go back and
forth between different types of seeding problems. We first give some
background on the different types of seeding methods considered here, we
then introduce the computational strategy based on transfer matrices, and
then delve into each specific seeding method. Finally, we provide a
statistical method to estimate the parameters of the problem from the
data.

@@

The term ``seed'' does not have exactly the same meaning according to
different authors. It can designate a part of the read, a part of the
genome, a particular sequence motif or a structured pattern of matches.
Also, the term does not always imply an exact match. For instance, the
algorithm PatternHunter~\cite{pmid11934743} uses ``spaced seeds''
that tolerate mismatches\footnote{Spaced seeds are covered
by patent US20070088510A1.}. To avoid potential confusion,
we will adopt the following convention.

@@

The index is usually a hash table or a variant of the so-called
FM-index~\cite{pmid20460430}. Hash tables are typically used to index
$k$-mers, which makes them useful to search seeds of set size $k$,
(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
algorithms). In contrast, the FM-index has no set size so it can be used
to search seeds of different lengths~\cite{ferragina2000opportunistic,
ferragina2005indexing}. It is a compact data structure based on the suffix
array~\cite{manber1993suffix} and on the Burrows-Wheeler
transform~\cite{burrows1994block}, emulating a suffix trie with a much
smaller memory footprint. Queries in the FM-index are performed with an
algorithm known as the \emph{backward
search}~\cite{ferragina2000opportunistic}.

Other methods can be efficient (\textit{e.g.}, running a bisection on the
suffix array~\cite{dobin2013star}) but the FM-index is currently the most
popular choice for seeding methods. For MEM seeds described below, it is
even the only practical option~\cite{pmid24336412, pmid25399029,
pmid23349213, pmid19389736}, so we implicitly assume that seeding is
performed by running the backward search. This will only have practical
consequences in section~\ref{sec:est}, where we propose a method to
estimate the parameters that are required to compute the probabilities of
interest. For the rest, we simply assume that seeds are known at all times
without ambiguity because this problem has several practical solutions.

@@

Fig.~\ref{fig:exact_seed_example} shows the example of a read produced
by an imperfect sequencer. The sequenced DNA fragment comes from a region
of the genome that has three duplicates (there are four homologous
sequences in total), and the read has three miscalled nucleotides
(indicated by stars).

@@

The exact seeds with $\gamma=7$ are shown above the sequence of the read.
Observe that erroneous nucleotides can belong to exact seeds because they
sometimes match one of the duplicates. For instance, the first sequencing
error matches all the duplicates and belongs to a seed for the first one.
However, erroneous nucleotides that are mismatches for \emph{all} the
duplicates cannot belong to a seed. This is the case of the second
sequencing error in this example, where there is a local deficit of
seeds.

Thie middle of the read shows some clutter where consecutive seeds
match consecutive sequences at the same location. This is typical for
exact seeds and constitutes a nuisance for the implementation. Indeed, it
is a waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive~\cite{pmid11934743} so it is used only for specific
applications. Nevertheless, it will be extremely useful for the
development of the present theory.

@@

Skip seeds have a fixed size $\gamma$, but unlike exact seeds they cannot
start at every nucleotide. Instead, a fixed amount of nucleotides is
skipped between every seed. This is a way to reduce the amount of
overlapping matches at the same location, at the cost of sensitivity. This
seeding heuristic is the core of Bowtie2~\cite{pmid22388286}, where seeds
have size $\gamma=16$ and are separated by 10 nucleotides (9 positions are
skipped). We will refer to seeds where $n$ nucleotides are skipped as
``skip-$n$ seeds''. For instance, Bowtie2 uses skip-9 seeds.

@@

Fig.~\ref{fig:skip_seed_example} shows what happens when exact seeds are
replaced by skip-2 seeds on the read of Fig.~\ref{fig:exact_seed_example}.
Here the size is still $\gamma=7$ but 1 nucleotide is skipped between
seeds. This amounts to removing every second seed. The consequence is that
there are fewer overlapping matches at the center of the read, but the
only seed for the second duplicate is lost. This is a rather positive
outcome because there is one off-target location fewer in the candidate
set, but the same might happen to the target.

@@

MEM seeds (where MEM stands for Maximal Exact Match) are somewhat harder
to define. Unlike exact seeds and skip seeds, their size is variable as we
are about to see. They are used in BWA-MEM~\cite{li2013aligning} where
they give good empirical results (but note that the actually seeding
scheme uses more complex Super Maximal Exact Matches). To describe MEM
seeds, let us first focus on MEMs.

@@

A \emph{MEM seed} is simply a MEM of minimum size $\gamma$.
Fig.~\ref{fig:MEM_example} shows what happens when exact seeds or skip
seeds are replaced by MEM seeds on the read of
Fig.~\ref{fig:exact_seed_example}. Observe that the clutter at the center
of the read has disappeared because consecutive matches are fused into a
single MEM seed.

@@

Compared to seeds of fixed size, MEM seeds have two counter-intuitive (and
undesirable) properties. The first is that there are cases where there
cannot be any good seed, even when changing the sensitivity parameter
$\gamma$. Fig.~\ref{fig:full_masking_example} shows such an example. Even
though there is a single sequencing error, the read has no MEM seed for
the target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location in this case (even though it is the best
location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
The read, the MEM seeds and the sequences are represented as in
Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two incorrect
locations at the bottom effectively hide the target, so it cannot
be discovered. This can occur even when the true location is the
best and when there is a single error on the read.}
\label{fig:full_masking_example}
\end{figure}

The second counter-intuitive property is that shortening a read can
sometimes generate a MEM seed for the target. Fig.~\ref{fig:short_vs_long}
shows an example of this case. There is no MEM seed for the target, but
there would be one if the read were two nucleotides shorter on the right
side. Indeed, in this case there would be a shared MEM seed matching the
first and the second locations (provided $\gamma \leq 12$).

@@

These examples show that MEM seeds can perform worse than seeds of fixed
size. MEM seeds yield fewer candidates and therefore speed up the mapping
process, but the question is at what cost for accuracy? The theory
developed here will allow us to compute the probability that a read has
no MEM seed for the target and thus that the true location is missed at
the seeding stage.

@@

Next, we assume that the target sequence has $N \geq 0$ duplicates, so
that there are $N$ off-target sequences. We further assume that
duplication happened instantaenously at some point in the past and that
all $N+1$ sequences diverge independently of each other at a constant
rate. In other words, we ignore the complications due to the genealogy of
the duplication events and we simply assume that at each nucleotide
position, any given duplicate is identical to the target with probability
$1-\mu$. If it is not, we assume that the duplicate sequence can be any of
one the remaining three nucleotides (\textit{i.e.} each is found with
probability $\mu/3$).

Note that read errors are always mismatches against the target (because we
assume that the target is the true sequence), and they match each
off-target sequence with probability $\mu/3$. In contrast, correct
nucleotides are always matches for the target, and they match each
off-target sequence with probability $1-\mu$.

Before going further, we also need to move a practical consideration out
of the way. Seeds can match any sequence of the genome, not just the
target or the duplicates. However, we will ignore matches in the rest of
the genome because such random matches are unlikely to cause a mapping
error when seeding is off target, contrary to matches in duplicates.
Neglecting those will greatly simplify the exposition of the theory
without loss of generality. We will explain in
section~\ref{sec:short_MEMs} how to deal with this practical case and and
how to identify those matches as off target. Until then, we will consider
that the target and the duplicates are the only sequences in the reference
genome.

@@

To give an example, assume that a particular symbol, say $\Downarrow$, has
a probability of occurence equal to $p$. The weighted generating function
of the word containing only this symbol is $pz$. The weight of the word is
its probability (here equal to $p$) and the size is its length in number
of symbols (here 1).

In this document we will focus on the weighted generating function $A(z)$
of the set $\mathcal{A}$ of reads that do not have on-target seeds
(\textit{i.e.} reads for which seeding is either off target or null). The
weight of a read is its probability of occurrence and the size $k$ is its
number of nucleotides. The coefficient $a_k$ is thus the proportion of
reads of size $k$ that do not have an on-target seed, which is the
quantity of interest.

@@

The analytic combinatorics framework relies on a strategy referred to as
the \emph{symbolice method}~\cite{sedgewick2013introduction}. The idea is
to combine simple objects into more complex objects. Each combinatorial
operation on the objects corresponds to a mathematical operation on their
weighted generating functions. One can thus obtain the weighted generating
function of complex objects, whose coefficients $a_k$ are the
probabilities of interest.

@@

The entry of $M(z) + M(z)^2 + M(z)^3 + \ldots = M(z) \cdot (I-M(z))^{-1}$
at coordinates $(i,j)$ is the weighted generating function of all the
reads that end with a segment of type $j$ and that can be appended to a
segment of type $i$. If segments of types $i$ and $j$ are exactly those
that can be placed at the beginning and at the end of the reads, then this
entry is the weighted generating function of all the reads without
on-target seed. The examples below will clarify the key steps of this
strategy.

A complete description of the strategy to compute seeding probabilities
with the symbolic method can be found in~\cite{filion2017analytic,
filion2018analytic}. The interested readers can also read more about
analytic combinatorics in the popular
textbooks~\cite{flajolet2009analytic, sedgewick2013introduction}.

@@

We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
\Downarrow\}$, where $\square$ represents a correct nucleotide,
$\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
appended after the last nucleotide to mark the end of the read.

This recoding allows us to partition the read in \emph{segments} defined
as a sequence of 0 or more symbols $\square$ followed by a symbol
$\Downarrow$. The last part of the read is called the tail and it is
terminated by the symbol $|$ instead of $\Downarrow$. The symbol $|$ does
not correspond to a nucleotide so it has size 0. Since the decomposition
is unique, we can view a read as a sequence of segments with a tail,
instead of a sequence of nucleotides.

Fig.~\ref{fig:simple} shows an example of decomposition in segments.
On-target seeds cannot contain sequencing errors, therefore they must be
completely embedded inside a segment or inside the tail. So the size of
the segments and of the tail indicates whether the read contains an
on-target seed or not.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_simple.pdf}
\caption{\textbf{The mismatch encoding.}
An example read is represented in the mismatch alphabet. The symbol
$\Downarrow$ represents a mismatch against the target (it is an erroneous
nucleotide) and the symbol $\square$ represents a match (it is a correct
nucleotide). The symbolic sequence of the target is represented below,
where an open square stands for a match and a closed square stands for a
mismatch.}
\label{fig:simple}
\end{figure}

The probability of occurrence of the symbol $\Downarrow$ is $p$, the error
rate of the sequencer, and the probability of occurrence of the symbol
$\square$ is thus $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that the weighted generating function of a segment
with $i$ symbols $\square$ is $(qz)^ipz$. The symbol $|$ has size 0, so
the weighted generating function of a tail segment with $i$ symbols
$\square$ is $(qz)^i$.

The key insight is that reads without on-target seed are exactly the reads
that are made of segments and tail with fewer than $\gamma$ symbols
$\square$. The weighted generating function of such segments is $pz \cdot
\big( 1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail is
$1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can be
encoded in a transfer matrix.

Since there are two kinds of objects, the segments and the tails, the
dimension of the transfer matrix is $2 \times 2$. A segment can be
followed by another segment or by the tail. The tail cannot be followed by
anything. The expression of the transfer matrix $M_0(z)$ is thus
\begin{equation*}
\begin{blockarray}{ccc}
   & \scriptstyle{\Downarrow} & \scriptstyle{|} \\
\begin{block}{c[cc]}
\scriptstyle{\Downarrow} & pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1}
\big)  & 1+qz+\ldots+(qz)^{\gamma-1} \\
\scriptstyle{|} & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

The types of objects are identified by their terminator, \textit{i.e.} by
their last symbol. Segments are identified by the terminator $\Downarrow$
and tails by the terminator $|$. The terminators are written in the
margins of the transfer matrix for readability.

We can compute the matrix $M_0(z) + M_0(z)^2 + \ldots = M_0(z) \cdot
(I-M_0(z))^{-1}$ and extract the entry of interest, which is the top right
term --- associated with terminators $\Downarrow$ and $|$. To see why,
observe that every read can be prepended by a segment and only by a
segment (not by a tail). Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow$ and that are terminated
by a tail, whose weighted generating function is the top right entry of
the matrix. It is easy to check that this term is equal to
\begin{equation}
\label{eq:F}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} = a_0 + a_1z + a_2z^2 +
\ldots
\end{equation}

>>>>

\title{On mapping errors and seeding heuristics for
DNA alignment}

@@

\begin{abstract}
% Statement of the problem.
The increasing throughput of DNA sequencing technologies demands faster
algorithms to analyze the data. The fate of most reads is to be mapped to
a reference sequence, typically a genome. Modern mappers rely on
heuristics to gain speed at a reasonable cost for accuracy. In the seeding
approach, short matches between the reads and the genome are used to
narrow the search to a set of candidate locations. Several seeding
variants used in modern mappers show good empirical performance but they
are difficult to calibrate or to optimize for lack of theoretical results.
% Summary of the results.
Here we develop a theory to estimate the probability that reads are
mismapped due to failures at the seeding step. We describe the properties
of simple exact seeds, skip-seeds and MEM seeds (Maximal Exact
Match).
% Innovation.
The major innovation is to use concepts from analytic combinatorics to
represent reads as abstract sequences, and to specify their generative
function to estimate the probabilities of interest.
% Durability.
We provide several algorithms, which combined together give a practicial
solution to the problem of calibrating seeding heuristic for short reads.
% Scope.
These results can improve current mapping algorithms and lay the
foundation of a general strategy to tackle sequence alignment problems.
\end{abstract}

@@

Suppose that we use and imperfect instrument to sequence a DNA fragment
drawn at random from a known genome; how can we identify the location of
the fragment in the genome?

We will refer to this question as the \emph{true location problem}.
It has applications in countless domains, like discovering
disease-causing mutations; diagnosing viral, bacterial or fungal
infections; diagnosing cancer and choosing appropriate therapies;
selecting breeds of interest in agriculture; tracing human migrations from
ancient bones; diagnosing genetic diseases at birth; finding compatible
graft donors \textit{etc}.

At first sight, whether a read can be mapped to its correct location seems
to depend only on its length and on the error rate of the sequencer.
However, there is another important factor to consider: most genomes
contain repeated sequences, \textit{i.e.} relatively large subsequences
that are present at more than one location. For instance, we cannot map
the fragment with certainty if there exists an exact copy of the sequence
somewhere else in the genome, because we cannot know which of the two
copies was sequenced.

This case is rare though. Repeated sequences are in general not identical
but merely homologous --- meaning that their similarity is unlikely to
occur by chance. So when the DNA fragment has one or more duplicates in
the genome, we can still map it, as long as we can distinguish the
duplicates. This in turn depends on their similarity with the target
fragment and on the error rate of the sequencer.

@@

We will refer to this task as the \emph{best location problem}. It amounts
to finding the optimal alignment between two sequences, and for this
reason has received substantial attention in bioinformatics. For
simplicity we will assume that the true location is always the best, and
we will often confound these two ideas.

@@

The slowness of exact alignment algorithms imposes a trade-off between
speed and accuracy: If the seeding step is set to produce many candidates,
it is likely to retrieve the candidate. The downside is that this will
take long. Conversely, if the seeding step is set to produce few
candidates, the process will run faster but the target is less likely to
be in the candidate set.

@@

The distinction is important because duplicates of the target are similar
to the read (due to their similarity to the target), so a failure of the
first kind is difficult to distinguish from a success. A failure of the
second kind is easier to recognize because in this case the best hit is
not similar to the read, as we will highlight in
section~\ref{sec:random_seeds}.

@@

If a seed-based mapping algorithm returns an incorrect location, then it
must be the case that the best location was never tested in the
post-filtration phase (otherwise it would have been found by the exact
alignment algorithm).

This observation is fundamental: in order to estimate the error rate of of
a mapping algorithm, we must know how often the associated seeding
heuristic is off target. For lack of theoretical framework, this
information is missing for nearly all existing seeding methods, meaning
that current mappers cannot be properly calibrated.

Our focus here is to develop a theory to estimate the probability that
seeding is off target, when using the most popular seeding heuristics for
short reads. Previous work pioneered a method to compute seeding
probabilities but it did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to control mapping errors. Other authors investigated
the reliability of mapping algorithms~\cite{pmid23872968}, but they
focused on the probability that random sequences have a single hit,
recognizing that solving the problem requires taking into consideration
the repeat structure of the genome. Here we address this problem in
general terms and we provide a practical way to estimate the mismapping
rate in complex genomes.

The rest of this document is organized to present the content in a
didactic manner. The order of the examples is meant to facilitate the
aquisition of the main concepts, meaning that we sometimes go back and
forth between different types of seeding problems. We first give some
background on the different types of seeding methods considered here, we
then introduce the computational strategy based on transfer matrices, and
then delve into each specific seeding method.

@@

The term ``seed'' changes meaning with different authors. It can designate
a part of the read, a part of the genome, a particular sequence motif or a
structured pattern of matches. Also, the term does not always imply an
exact match. For instance, the algorithm PatternHunter~\cite{pmid11934743}
uses ``spaced seeds'' that tolerate mismatches. To avoid potential
confusion, we will adopt the following convention.

@@

The index is usually a hash table or a variant of the so-called
FM-index~\cite{pmid20460430}. Hash tables are typically used to index
$k$-mers, which makes them useful to search seeds of set size $k$,
(see~\cite{pmid30346548} for a recent benchmark of $k$-mer hashing
algorithms). In contrast, the FM-index has no set size so it can be used
to search seeds of different lengths~\cite{ferragina2000opportunistic,
ferragina2005indexing}. It is a compact data structure based on the suffix
array~\cite{manber1993suffix} and on the Burrows-Wheeler
transform~\cite{burrows1994block}, emulating a suffix trie with a much
smaller memory footprint.

Other methods can be efficient (\textit{e.g.}, running a bisection on the
suffix array~\cite{dobin2013star}) but the FM-index is currently the most
popular choice for seeding methods. For MEM seeds described below, it is
even the only practical option~\cite{pmid24336412, pmid25399029,
pmid23349213, pmid19389736}. However, the detail if of little interest
here: we simply assume that seeds are known at all times without ambiguity
because this problem has several practical solutions.

@@

Fig.~\ref{fig:exact_seed_example} shows the example of a read produced
by an imperfect sequencer. The sequenced DNA fragment comes from a region
of the genome that has three duplicates and the read has three miscalled
nucleotides (indicated by stars).

@@

The exact seeds with $\gamma=7$ are shown above the sequence of the read.
Observe that erroneous nucleotides can belong to exact seeds because they
sometimes match one of the duplicates. For instance, the first sequencing
error matches all the duplicates and belongs to an off-target seed.
However, erroneous nucleotides that are mismatches for \emph{all} the
sequences cannot belong to a seed. This is the case of the second
sequencing error in this example, where there is a local deficit of seeds.

The middle of the read shows some clutter where consecutive seeds
match consecutive sequences at the same location. This is typical for
exact seeds and constitutes a nuisance for the implementation. Indeed, it
is a waste of computer resources to discover matches in sequences that are
already in the candidate set. In addition, this seeding method is not
particularly sensitive compared to spaced seed~\cite{pmid11934743} so it
is used only in few specific applications. Nevertheless, it will be
extremely useful for the development of the present theory.

@@

Skip seeds have a fixed size $\gamma$, but unlike exact seeds they cannot
start at every nucleotide. Instead, a certain amount of nucleotides is
skipped between every seed. This is a way to reduce the overlapping
matches at the same location, at the cost of sensitivity. This seeding
heuristic is the core of Bowtie2~\cite{pmid22388286}, where seeds have
size $\gamma=16$ and are separated by 10 nucleotides (9 positions are
skipped). We will refer to seeds where $n$ nucleotides are skipped as
``skip-$n$ seeds''. For instance, Bowtie2 uses skip-9 seeds.

@@

Fig.~\ref{fig:skip_seed_example} shows what happens when exact seeds are
replaced by skip-1 seeds on the read of Fig.~\ref{fig:exact_seed_example}.
Here the size is still $\gamma=7$ but 1 nucleotide is skipped between
seeds. This amounts to removing every second seed. The consequence is that
there are fewer overlapping matches at the center of the read, but the
only seed for the second duplicate is lost. This is a rather positive
outcome because there is one off-target location fewer in the candidate
set, but the same might happen to the target.

@@

MEM seeds (where MEM stands for Maximal Exact Match) are somewhat harder
to define. Unlike exact seeds and skip seeds, their size is variable. They
are used in BWA-MEM~\cite{li2013aligning} where they give good empirical
results. To describe MEM seeds, let us first focus on MEMs.

@@

A \emph{MEM seed} is simply a MEM of size size $\gamma$ or greater.
Fig.~\ref{fig:MEM_example} shows what happens when using MEM seeds on the
read of Fig.~\ref{fig:exact_seed_example}. Observe that the clutter at the
center of the read has disappeared because consecutive matches are fused
into a few MEM seeds.

@@

Compared to seeds of fixed size, MEM seeds have two counter-intuitive
properties. The first is that there are cases where there cannot be any
on-target seed, even when changing the sensitivity parameter $\gamma$.
Fig.~\ref{fig:full_masking_example} shows such an example. Even though
there is a single sequencing error, the read has no MEM seed for the
target. Lowering $\gamma$ does not change this, so there is no way to
discover the true location in this case (even though it is the best
location).

\begin{figure}[h]
\centering
\includegraphics[scale=1]{full_masking_example.pdf}
\caption{\textbf{Issues with MEM seeds: inaccessible targets.}
The read, the MEM seeds and the sequences are represented as in
Fig.~\ref{fig:MEM_example}. The MEM seeds matching the two duplicates
at the bottom effectively hide the target, so it cannot be discovered.
This can occur even when the true location is the best and when there is a
single error on the read.}
\label{fig:full_masking_example}
\end{figure}

The second counter-intuitive property is that shortening a read can
sometimes generate a MEM seed for the target. Fig.~\ref{fig:short_vs_long}
shows an example of this case. There is no MEM seed for the target, but
there would be if the read were two nucleotides shorter on the right
side. Indeed, in this case there would be a shared MEM seed matching the
the target and the first duplicate (provided $\gamma \leq 12$).

@@

These examples show that MEM seeds can perform worse than seeds of fixed
size. MEM seeds yield fewer candidates and therefore speed up the mapping
process, but the question is at what cost? The theory developed here will
allow us to compute the probability that a read has no MEM seed for the
target and thus that the true location is missed at the seeding stage.

@@

Next, we assume that the target sequence has $N \geq 0$ duplicates, so
that there are $N$ off-target sequences. We further assume that
duplication happened instantaneously at some point in the past and that
all $N+1$ sequences diverge independently of each other at a constant
rate. In other words, we ignore the complications due to the genealogy of
the duplication events and we simply assume that at each nucleotide
position, any given duplicate is identical to the target with probability
$1-\mu$. If it is not, we assume that the duplicate sequence can be any of
one the remaining three nucleotides (\textit{i.e.} each is found with
probability $\mu/3$).

Note that read errors are always mismatches against the target (because we
assume that the target is the true sequence), and they match each
duplicate with probability $\mu/3$. Correct nucleotides are always matches
for the target, and they match each duplicate with probability $1-\mu$.

Before going further, we also need to move a practical consideration out
of the way. Seeds can match any sequence of the genome, not just the
target or the duplicates. However, we will ignore matches in the rest of
the genome because such random matches are unlikely to cause a mapping
error when seeding is off target, contrary to matches in duplicates.
Neglecting those will greatly simplify the exposition of the theory
without loss of generality. We will explain in
section~\ref{sec:random_seeds} how to deal with this practical case and
and how to identify those matches as off target. Until then, we will
consider that the target and the duplicates are the only sequences in the
reference genome.

@@

To give an example, assume that a particular symbol, say $\Downarrow$, has
a probability of occurence equal to $p$. The weighted generating function
words containing only this symbol is $pz$. The weight of the word is
its probability (here equal to $p$) and the size is its length in number
of symbols (here 1).

In this document we will focus on the weighted generating function $A(z)$
of the set $\mathcal{A}$ of reads that do not have on-target seeds
(\textit{i.e.} reads for which seeding is either null or off target). The
weight of a read is its probability of occurrence and the size $k$ is its
number of nucleotides. The coefficient $a_k$ is thus the proportion of
reads of size $k$ that do not have an on-target seed, which is the
quantity of interest.

@@

The analytic combinatorics framework relies on a strategy referred to as
the \emph{symbolic method}~\cite{sedgewick2013introduction}. The idea is
to combine simple objects into more complex objects. Each combinatorial
operation on the objects corresponds to a mathematical operation on their
weighted generating functions. One can thus obtain the weighted generating
function of complex objects, whose coefficients $a_k$ are the
probabilities of interest.

@@

From the basic operations on weighted generating functions, the entry of
$M(z) + M(z)^2 + M(z)^3 + \ldots = M(z) \cdot (I-M(z))^{-1}$ at
coordinates $(i,j)$ is the weighted generating function of all the reads
that end with a segment of type $j$ and that can be appended to a segment
of type $i$. The examples below will clarify the key steps of this
strategy.

A complete description of how to compute seeding probabilities with the
symbolic method can be found in~\cite{filion2017analytic,
filion2018analytic}. The interested readers can also find more about
analytic combinatorics in the popular
textbooks~\cite{flajolet2009analytic, sedgewick2013introduction}.

@@

We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
\Downarrow\}$, where $\square$ represents a correct nucleotide,
$\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
appended after the last nucleotide to mark the end of the read. It is not
associated to a nucleotide and therefore has size 0.

This recoding allows us to partition the read in am important way.

\begin{definition}
\label{def:seg}
Every symbol of the alphabet that is different from $\square$ is called a
terminator. A segment is a sequence of 0 or more symbols $\square$
followed by a terminator. The last segment of the read is terminated by
the symbol $|$ and is called the tail.
\end{definition}

Since the decomposition of the read in segments is unique, we can view a
read as a sequence of segments with a tail, instead of a sequence of
nucleotides. Fig.~\ref{fig:simple} shows an example of decomposition in
segments. On-target seeds cannot contain sequencing errors, therefore they
must be completely embedded inside a segment. So the sizes of the segments
indicate whether the read contains an on-target seed or not.

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_simple.pdf}
\caption{\textbf{The mismatch encoding.}
An example read is represented in the mismatch alphabet. The symbol
$\Downarrow$ represents a mismatch against the target (it is an erroneous
nucleotide) and the symbol $\square$ represents a match (it is a correct
nucleotide). The symbol $|$ is appended to the end of the read. The
symbolic sequence of the target is represented below, where an open square
stands for a match and a closed square stands for a mismatch.}
\label{fig:simple}
\end{figure}

The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
rate of the sequencer) and the probability of occurrence of the symbol
$\square$ is thus $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that the weighted generating function of a segment
with $i$ symbols $\square$ is $(qz)^ipz$. The symbol $|$ has size 0, so
the weighted generating function of a tail segment with $i$ symbols
$\square$ is $(qz)^i$.

The key insight is that reads without on-target seed are exactly the reads
that are made of segments with fewer than $\gamma$ symbols $\square$. The
weighted generating function of such segments is $pz \cdot \big(
1+qz+\ldots+(qz)^{\gamma-1} \big)$, and that of the tail is
$1+qz+\ldots+(qz)^{\gamma-1}$. This gives a construction plan that can be
encoded in a transfer matrix.

There are two kinds of objects: the $\Downarrow$ segments and the tails,
so the dimension of the transfer matrix is $2 \times 2$. A $\Downarrow$
segment can be followed by another $\Downarrow$ segment or by the tail.
The tail cannot be followed by anything. The expression of the transfer
matrix $M_0(z)$ is thus
\begin{equation*}
\begin{blockarray}{ccc}
   & \scriptstyle{\Downarrow} & \scriptstyle{|} \\
\begin{block}{c[cc]}
\scriptstyle{\Downarrow} & pz \cdot \big( 1+qz+\ldots+(qz)^{\gamma-1}
\big)  & 1+qz+\ldots+(qz)^{\gamma-1} \\
\scriptstyle{|} & 0 & 0 \\
\end{block}
\end{blockarray}.
\end{equation*}

In the representation above, the different types of objects are identified
by their terminator. The terminators are written in the margins of the
transfer matrix for readability.

We can compute the matrix $M_0(z) + M_0(z)^2 + \ldots = M_0(z) \cdot
(I-M_0(z))^{-1}$ and extract the entry of interest, which is the top right
term --- associated with terminators $\Downarrow$ and $|$. To see why,
observe that every read can be prepended by $\Downarrow$ segments and only
by those (not by a tail). Thus, reads are precisely the sequences of
segments that can follow the symbol $\Downarrow$ and that are terminated
by a tail, whose weighted generating function is the top right entry of
the matrix. It is easy to check that this term is equal to
\begin{equation}
\label{eq:F}
\frac{1+qz+\ldots+(qz)^{\gamma-1}}
  {1-pz \big(1+qz+\ldots+(qz)^{\gamma-1} \big)} = a_0 + a_1z + a_2z^2 +
\ldots
\end{equation}
